Type Classes for Mathematics in Type Theory

### 2010-07-01 jar.tex
Clarify use of naturals_to_semiring, bigops.

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {σ}{{$\upsigma$}}1
  {η}{{$\upeta$}}1
  {φ}{{$\upphi$}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {⇛}{{$\Rrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi$ }}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot$}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
  {~}{{$\sim$}}1
}

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then lifted to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using the fact that \lstinline|naturals_to_semiring| is an involution.

For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour. As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations. Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us transfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings. We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quoting}.

@@

  Class Extend := extend: Π {x y} `{SemiGroupOp y} `{MonoidUnit y},
    (x → y) → (seq x → y).
  Class Singleton := singleton: Π x, x → seq x.
\end{lstlisting}
With these, we can define the predicate class for sequences:
\begin{lstlisting}
  Class Sequence
   `{Π a, MonoidUnit (seq a)} `{Π a, SemiGroupOp (seq a)}
   `{Π a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
\end{lstlisting}
%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
%    ; ... AltAdjunction stated using the bundled ... }.
On top of this interface we can build theory about typical list operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera. We can also generically define `bigops' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.

>>>>

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {σ}{{$\upsigma$}}1
  {η}{{$\upeta$}}1
  {φ}{{$\upphi$}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {⇛}{{$\Longrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi$ }}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot$}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
  {~}{{$\sim$}}1
}

@@

Since initial objects in categories are isomorphic, we can easily derive that \lstinline|naturals_to_semiring| gives isomorphisms between different \lstinline|Naturals| implementations:
\begin{lstlisting}
  Lemma iso_naturals `{Naturals A} `{Naturals B}:
    Π a: A, naturals_to_semiring B A (naturals_to_semiring A B a) = a.
\end{lstlisting}
This is very useful, because some properties of and operations on naturals are more easily proved, respectively defined, for concrete implementations (such as \lstinline|nat|) and then `lifted' to the abstract \lstinline|Naturals| interface so that they work for arbitrary implementations. For example, while showing decidability for an arbitrary \lstinline|Naturals| implementation directly is tricky, it is very easy to show decidability for \lstinline|nat|. Using \lstinline|iso_naturals|, the latter can be very straightforwardly used to implement the former.

To lift properties such as injectivity of partially applied addition and multiplication from \lstinline|nat| to arbitrary \lstinline|Naturals| implementations, we take a longer detour. As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations. Hence, we can transfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings. We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in Section~\ref{quoting}.

@@

  Class Extend := extend: Π {x y} `{SemiGroupOp y} `{MonoidUnit y},
    (x → y) → (seq x → y).
  Class Singleton := singleton: Π x, x → seq x.
\end{lstlisting}
With these, we can define the predicate class for sequences:
\begin{lstlisting}
  Class Sequence
   `{Π a, MonoidUnit (seq a)} `{Π a, SemiGroupOp (seq a)}
   `{Π a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
\end{lstlisting}
%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
%    ; ... AltAdjunction stated using the bundled ... }.
On top of this interface we can build theory about typical sequence operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera. We can also generically define `big operators' for sums ($\sum$) and products ($\prod$) of sequences, and easily show properties like distributivity, all without ever mentioning cons lists.
